Natural regeneration of a tropical timber plantation

### 2022-04-14 talk/talk.Rmd
made fig3
update figs1,2, map next

---
title: 
  "Natural regeneration of a tropical timber plantation"
author: 
  "Nicholas Medina"
output: 
  # slidy_presentation
  # oislides_presentation
  # powerpoint_presentation
  html_notebook
  # html_document
  # pdf_document #no
---

@@

Outline
===

@@

I--Global forests are regenerating, but how?
===

@@

- Regeneration dynamics should be better understood

I--Regeneration is quick and inconsistent
===

- Biomass recovers 90% in 66 yrs *(Poorter et al 2016 PNAS)*

- Taxonomic richness recovers quickly ()

- Community composition may take centuries to recover, if at all *(Norder et al 2010 PNAS)*

I--Edges shape regeneration
===

- 70% forests are <1 km from a habitat edge *(Haddad et al 2015 Science)*

- Edges effects tend to reduce biodiversity by 13 - 75% via ... *(")*

    - stronger in small and isolated fragments

I--Edge effects may differ in wet tropical plantations
===

@@

- Wet tropical forest regeneration tends to be light-limited, based on tree functional traits *(Poorter et al 2021)*

@@

I--This study
===

@@

II--Study site
===

@@

II--Census design
===

@@

II--Census measurements
===

@@

II--Statistical analysis
===

@@

- Traits obtained from literature and compared to TRY database *(Kattge et al 2020)*

- Computations done with R `getRversion()` depending on *BIOMASS* `packageVersion("BIOMASS")`, 
*vegan* `packageVersion("vegan")`, and 
*tidyverse* `packageVersion("tidyverse")` packages

III--Stand regeneration
===

@@

III--Composition regeneration
===

@@

        - Vochysia tended to decrease

        - Ficus decreased

@@

III--Functional regeneration
===

-

-

-

@@

IV--Hypothesis check
===

@@

@@

IV--Processes underlying patterns
===

- not shade...

- dispersal...

-

IV--Compare to similar studies
===

- aligns well with major recent papers

-

-

IV--Future studies
===

@@

>>>>

---
title: 
  "Natural regeneration of a tropical timber plantation"
author: 
  "Nicholas Medina"
output: 
  # slidy_presentation
  # oislides_presentation
  # powerpoint_presentation
  html_notebook
  # pdf_document #no
  # word_document
bibliography: ../../library.bib
---

@@

# Outline

@@

# I--Global forests are regenerating, but how?

@@

- Regeneration dynamics (e.g. associated biodiversity) should be better understood to optimize sustainable management and service provision

# I--Forest regeneration is overall both quick and inconsistent

- Biomass recovers 90% in 66 yrs *[@poorter16]*

- Taxonomic richness recovers quickly *[@rozendaal19]*

- Community composition may take centuries to recover, if at all *[@norden15]*

# I--Habitat edges shape regeneration

- 70% forests are <1 km from a habitat edge *[@haddad15]*

- Edge effects reduce biodiversity 13 - 75%

    - more so in small and isolated fragments

# I--Edge effects may differ in wet tropical plantations

@@

- Wet tropical forest regeneration tends to be light-limited, based on tree functional traits *[@poorter21]*

@@

# I--This study

@@

# II--Study site

@@

# II--Census design

@@

# II--Census measurements

@@

# II--Statistical analysis

@@

- Traits obtained from literature and compared to TRY database *[@kattge19]*

- Computations done with R `r getRversion()` depending on *BIOMASS* `r packageVersion("BIOMASS")`, 
*vegan* `r packageVersion("vegan")`, and 
*tidyverse* `r packageVersion("tidyverse")` packages

# III--Stand regeneration

@@

# III--Composition regeneration

@@

        - *Vochysia* tended to decrease

        - *Ficus* decreased

@@

# III--Functional regeneration

- Away from forest edge--

    - Biomass of taxa often found in both successional stages decreased

    - Biomass based on dispersal mode remained constant

@@

# IV--Hypothesis check

@@

        - Composition tended to change

@@

# IV--Processes underlying patterns

- Not shade, perhaps given high absolute value and/or evenness of original timber planting

- Forest edge may associate with faster succession, given drop in generalist tree taxa with distance

- No signal of dispersal mode detected

# IV--Compare to similar studies

- Aligns well with several recent studies suggesting that--

    - Tree diversity regenerates quicker than taxonomic composition *[@poorter16]*

    - Habitat edges lower diversity and generally slow regeneration, like via less seed consumption *[@hohlenwerger22]*

    - Taxa indiscriminate to successional stage are minimally affected by disturbance, like from edges *[@bongers09]*

# IV--Future studies

@@

# Acknowledgements

- James Fifer for initial review

- Lehigh group Don Morris and undergrads for field help

- Brandeis University for funding

\pagebreak

# References

### 2022-04-14 talk/talk.tex
update figs1,2, map next

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{Natural regeneration of a tropical timber plantation}
\author{Nicholas Medina}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Natural regeneration of a tropical timber plantation},
  pdfauthor={Nicholas Medina},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\begin{itemize}
\item
  \textasciitilde Half of forests are tropical, and over half of all
  forests globally are secondary \emph{(FAO 2020)}
\item
  Services like timber and biodiversity depend on natural and managed
  regeneration patterns
\item
  Regeneration dynamics should be better understood
\end{itemize}

\hypertarget{iregeneration-is-quick-and-inconsistent}{%
\section{I--Regeneration is quick and
inconsistent}\label{iregeneration-is-quick-and-inconsistent}}

\begin{itemize}
\item
  Biomass recovers 90\% in 66 yrs \emph{(Poorter et al 2016 PNAS)}
\item
  Taxonomic richness recovers quickly ()
\item
  Community composition may take centuries to recover, if at all
  \emph{(Norder et al 2010 PNAS)}
\end{itemize}

\hypertarget{iedges-shape-regeneration}{%
\section{I--Edges shape regeneration}\label{iedges-shape-regeneration}}

\begin{itemize}
\item
  70\% forests are \textless1 km from a habitat edge \emph{(Haddad et al
  2015 Science)}
\item
  Edges effects tend to reduce biodiversity by 13 - 75\% via \ldots{}
  \emph{(``)}

  \begin{itemize}
  \tightlist
  \item
    stronger in small and isolated fragments
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Often dense with fast-growing, shade-intolerant taxa
  \end{itemize}
\item
  Wet tropical forest regeneration tends to be light-limited, based on
  tree functional traits \emph{(Poorter et al 2021)}
\item
  Existing plantation demography and shade may skew/buffer edge effects
  toward net positive instead of negative
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Trend lines follow median values of all plots in distance stratum
  \end{itemize}
\item
  Traits obtained from literature and compared to TRY database
  \emph{(Kattge et al 2020)}
\item
  Computations done with R \texttt{getRversion()} depending on
  \emph{BIOMASS} \texttt{packageVersion("BIOMASS")}, \emph{vegan}
  \texttt{packageVersion("vegan")}, and \emph{tidyverse}
  \texttt{packageVersion("tidyverse")} packages
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
\item
\item
\end{itemize}

@@

    \begin{itemize}
    \item
      No trends in biomass
    \item
      Diversity decreased
    \end{itemize}
  \end{itemize}
\item
  Forest edge associated with tree diversity
\end{itemize}

@@

\begin{itemize}
\item
  not shade\ldots{}
\item
  dispersal\ldots{}
\item
\end{itemize}

@@

\begin{itemize}
\item
  aligns well with major recent papers
\item
\item
\end{itemize}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Natural regeneration of a tropical timber plantation},
  pdfauthor={Nicholas Medina},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Natural regeneration of a tropical timber plantation}
\author{Nicholas Medina}
\date{}

@@

\begin{itemize}
\item
  \textasciitilde Half of forests are tropical, and over half of all
  forests globally are secondary \emph{(FAO 2020)}
\item
  Services like timber and biodiversity depend on natural and managed
  regeneration patterns
\item
  Regeneration dynamics (e.g.~associated biodiversity) should be better
  understood to optimize sustainable management and service provision
\end{itemize}

\hypertarget{iforest-regeneration-is-overall-both-quick-and-inconsistent}{%
\section{I--Forest regeneration is overall both quick and
inconsistent}\label{iforest-regeneration-is-overall-both-quick-and-inconsistent}}

\begin{itemize}
\item
  Biomass recovers 90\% in 66 yrs \emph{(Poorter et al. 2016)}
\item
  Taxonomic richness recovers quickly \emph{(Rozendaal et al. 2019)}
\item
  Community composition may take centuries to recover, if at all
  \emph{(Norden et al. 2015)}
\end{itemize}

\hypertarget{ihabitat-edges-shape-regeneration}{%
\section{I--Habitat edges shape
regeneration}\label{ihabitat-edges-shape-regeneration}}

\begin{itemize}
\item
  70\% forests are \textless1 km from a habitat edge \emph{(Haddad et
  al. 2015)}
\item
  Edge effects reduce biodiversity 13 - 75\%

  \begin{itemize}
  \tightlist
  \item
    more so in small and isolated fragments
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Often dense with fast-growing, shade-intolerant taxa
  \end{itemize}
\item
  Wet tropical forest regeneration tends to be light-limited, based on
  tree functional traits \emph{(Poorter et al. 2021)}
\item
  Existing plantation demography and shade may skew/buffer edge effects
  toward net positive instead of negative
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Trend lines follow median values of all plots in distance stratum
  \end{itemize}
\item
  Traits obtained from literature and compared to TRY database
  \emph{(Kattge and Str 2019)}
\item
  Computations done with R 4.1.3 depending on \emph{BIOMASS} 2.1.7,
  \emph{vegan} 2.5.7, and \emph{tidyverse} 1.3.1 packages
\end{itemize}

@@

\begin{itemize}
\item
  Away from forest edge--

  \begin{itemize}
  \item
    Biomass of taxa often found in both successional stages decreased
  \item
    Biomass based on dispersal mode remained constant
  \end{itemize}
\end{itemize}

@@

    \begin{itemize}
    \item
      No trends in biomass
    \item
      Diversity decreased
    \item
      Composition tended to change
    \end{itemize}
  \end{itemize}
\item
  Forest edge associated with tree diversity
\end{itemize}

@@

\begin{itemize}
\item
  Not shade, perhaps given high absolute value and/or evenness of
  original timber planting
\item
  Forest edge may associate with faster succession, given drop in
  generalist tree taxa with distance
\item
  No signal of dispersal mode detected
\end{itemize}

@@

\begin{itemize}
\item
  Aligns well with several recent studies suggesting that--

  \begin{itemize}
  \item
    Tree diversity regenerates quicker than taxonomic composition
    \emph{(Poorter et al. 2016)}
  \item
    Habitat edges lower diversity and generally slow regeneration, like
    via less seed consumption \emph{(Hohlenwerger, Tambosi, and Metzger
    2022)}
  \item
    Taxa indiscriminate to successional stage are minimally affected by
    disturbance, like from edges \emph{(Bongers et al. 2009)}
  \end{itemize}
\end{itemize}

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bongers09}{}}%
Bongers, Frans, Lourens Poorter, William D. Hawthorne, and Douglas
Sheil. 2009. {``The Intermediate Disturbance Hypothesis Applies to
Tropical Forests, but Disturbance Contributes Little to Tree
Diversity.''} \emph{Ecology Letters} 12 (8): 798--805.
\url{https://doi.org/10.1111/j.1461-0248.2009.01329.x}.

\leavevmode\vadjust pre{\hypertarget{ref-haddad15}{}}%
Haddad, Nick M., Lars A. Brudvig, Jean Clobert, Kendi F. Davies, Andrew
Gonzalez, Robert D. Holt, Thomas E. Lovejoy, et al. 2015. {``Habitat
Fragmentation and Its Lasting Impact on {Earth}'s Ecosystems.''}
\emph{Science Advances} 1 (2): e1500052.
\url{https://doi.org/10.1126/sciadv.1500052}.

\leavevmode\vadjust pre{\hypertarget{ref-hohlenwerger22}{}}%
Hohlenwerger, Camila, Leandro Reverberi Tambosi, and Jean Paul Metzger.
2022. {``Forest Cover and Proximity to Forest Affect Predation by
Natural Enemies in Pasture and Coffee Plantations Differently.''}
\emph{Agriculture, Ecosystems \& Environment} 333 (August): 107958.
\url{https://doi.org/10.1016/j.agee.2022.107958}.

\leavevmode\vadjust pre{\hypertarget{ref-kattge19}{}}%
Kattge, Jens, and Hans Knöll Str. 2019. {``{TRY} Plant Trait Database
\textendash{} Enhanced Coverage and Open Access,''} 70.

\leavevmode\vadjust pre{\hypertarget{ref-norden15}{}}%
Norden, Natalia, Héctor A. Angarita, Frans Bongers, Miguel
Martínez-Ramos, Iñigo Granzow-de la Cerda, Michiel van Breugel, Edwin
Lebrija-Trejos, et al. 2015. {``Successional Dynamics in {Neotropical}
Forests Are as Uncertain as They Are Predictable.''} \emph{Proceedings
of the National Academy of Sciences} 112 (26): 8013--18.
\url{https://doi.org/10.1073/pnas.1500403112}.

\leavevmode\vadjust pre{\hypertarget{ref-poorter16}{}}%
Poorter, Lourens, Frans Bongers, T. Mitchell Aide, Angélica M. Almeyda
Zambrano, Patricia Balvanera, Justin M. Becknell, Vanessa Boukili, et
al. 2016. {``Biomass Resilience of {Neotropical} Secondary Forests.''}
\emph{Nature} 530 (7589): 211--14.
\url{https://doi.org/10.1038/nature16512}.

\leavevmode\vadjust pre{\hypertarget{ref-poorter21}{}}%
Poorter, Lourens, Dylan Craven, Catarina C. Jakovac, Masha T. van der
Sande, Lucy Amissah, Frans Bongers, Robin L. Chazdon, et al. 2021.
{``Multidimensional Tropical Forest Recovery.''} \emph{Science} 374
(6573): 1370--76. \url{https://doi.org/10.1126/science.abh3629}.

\leavevmode\vadjust pre{\hypertarget{ref-rozendaal19}{}}%
Rozendaal, Danaë M. A., Frans Bongers, T. Mitchell Aide, Esteban
Alvarez-Dávila, Nataly Ascarrunz, Patricia Balvanera, Justin M.
Becknell, et al. 2019. {``Biodiversity Recovery of {Neotropical}
Secondary Forests.''} \emph{Science Advances} 5 (3).
\url{https://doi.org/10.1126/sciadv.aau3114}.

\end{CSLReferences}
